# dplyr vs pandas

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r lib-r}
library(dplyr)
```

```{python lib-py}
import pandas as pd
```

- `select()` picks variables based on their names.


- `mutate()` adds new variables that are functions of existing variables


- `filter()` picks cases based on their values.

- `summarise()` reduces multiple values down to a single summary.

- `arrange()` changes the ordering of the rows.

See <https://dplyr.tidyverse.org/> for more details.

We will use the following toy data to apply the verbs. 

```{r df0-r, include = FALSE}
df <- tibble(
  name = c("Barney", "Ted", "Marshall",
           "Lilly","Robin"),
  gender = c("Male", "Male","Male",
             "Female", "Female"),
  grade = c(10, 11, 13, 12, 14)
)
df
```

```{r, echo = FALSE}
knitr::kable(df)
```

## Create Dataframe

<table class='table'>
<tr> <th>dplyr</th> <th>pandas</th> <tr>
<tr>
<td>
```{r df-r}
df <- tibble(
  name = c("Barney", "Ted", "Marshall",
           "Lilly","Robin"),
  gender = c("Male", "Male","Male",
             "Female", "Female"),
  grade = c(10, 11, 13, 12, 14)
)
df
```

</td>
<td>
```{python df-py}
df = pd.DataFrame({
  'name':["Barney", "Ted", "Marshall",
          "Lilly", "Robin"],
  'gender':["Male", "Male","Male", 
            "Female", "Female"],
  'grade':[10, 11, 13, 12, 14] 
})
df
```
</td>
<tr>
</table>


## Data Structure

<table class='table'>
<tr> <th>dplyr</th> <th>pandas</th> <tr>
<tr>
<td>

```{r glimpse-r}
glimpse(df)
```
</td>
<td>
```{python dtypes-py}
df.dtypes
df.shape
df.info()
```
</td>
<tr>
</table>



## select()

<span style="color:blue">*Select the variables `name` and `grade`.
*</span>

<table class='table'>
<tr> <th>dplyr</th> <th>pandas</th> <tr>
<tr>
<td>

```{r select-r}
df %>% 
  select(name, grade)

```

</td>
<td>
```{python select-py}
df[['name', 'grade']]

```

```{python}
# or
df.drop(columns = ['grade'])
```

```{python}
# or
df.drop(['grade'], axis = 1)
```

</td>
<tr>
</table>



