# Grammar: dplyr vs pandas & numpy

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r lib-r, echo = FALSE}
library(dplyr)
library(reticulate)
```

```{python lib-py, echo = FALSE}
import pandas as pd
import numpy as np
```



We will use the five [dplyr verbs](https://dplyr.tidyverse.org/) (also [pandas' guide](https://pandas.pydata.org/pandas-docs/stable/getting_started/comparison/comparison_with_r.html)) for comparison

- `select()` picks variables based on their names.


- `mutate()` adds new variables that are functions of existing variables


- `filter()` picks cases based on their values.

- `summarise()` reduces multiple values down to a single summary.

- `arrange()` changes the ordering of the rows.


and use the following toy data to apply the verbs.



|name     |gender | grade|
|:--------|:------|-----:|
|Barney   |Male   |    10|
|Ted      |Male   |    11|
|Marshall |Male   |    13|
|Lilly    |Female |    12|
|Robin    |Female |    14|




<span style="color:blue">**Create Toy Data**</span>


<table class='table'>
<tr> <th>dplyr</th> <th>pandas</th> <tr>
<tr>
<td>
```{r df-r}
df <- tibble(
  name = c("Barney", "Ted", "Marshall",
           "Lilly","Robin"),
  gender = c("Male", "Male","Male",
             "Female", "Female"),
  grade = c(10, 11, 13, 12, 14)
)
df
```

</td>
<td>
```{python df-py}
df = pd.DataFrame({
  'name':["Barney", "Ted", "Marshall",
          "Lilly", "Robin"],
  'gender':["Male", "Male","Male", 
            "Female", "Female"],
  'grade':[10, 11, 13, 12, 14] 
})
df
```
</td>
<tr>
</table>



<span style="color:blue">**Check Data Structure**</span>

<table class='table'>
<tr> <th>dplyr</th> <th>pandas</th> <tr>
<tr>
<td>

```{r glimpse-r}
glimpse(df)
```
</td>
<td>
```{python dtypes-py}
df.dtypes
df.shape
df.info()
```
</td>
<tr>
</table>



## `select()`

### *Example: Pick the variables `name` and `grade`.*</span>

<table class='table'>
<tr> <th>dplyr</th> <th>pandas</th> <tr>
<tr>
<td>

```{r select-r}
df %>% 
  select(name, grade)

```

</td>
<td>
```{python select-py}
df[['name', 'grade']]

```

```{python}
# or
df.drop(columns = ['gender'])
```

```{python}
# or
df.drop(['gender'], axis = 1)
```

**Using positions of columns:**

```{python}
df[df.columns[[0,2]]]
```

```{python}
df.iloc[:, [0,2]]
```


</td>
<tr>
</table>

## `mutate()`

### *Example: Generate a variable `grade_p`, expressing grade out of 100.*</span>

<table class='table'>
<tr> <th>dplyr</th> <th>pandas</th> <tr>
<tr>
<td>

```{r mutate-r}
df %>% 
  mutate(grade_p = grade/20*100)

```

</td>
<td>
```{python mutate-py}
df['grade_p'] = df['grade']/20*100
df
```
```{python}
# now drop the newly created variable
df.drop(columns = 'grade_p', inplace = True)
```

</td>
<tr>
</table>

## `filter()`

### *Example: Keep if the student is Barney or female.*

<table class='table'>
<tr> <th>dplyr</th> <th>pandas</th> <tr>
<tr>
<td>

```{r filter-r}
df %>% 
  filter(name == "Barney"|
         gender == "Female")

```

</td>
<td>
```{python filter-py}
# similar to base R
df[(df["name"] == "Barney") | 
   (df["gender"] == "Female")]
```

```{python filter2-py}
# query with ''; need to use "" for conditions
df.query('name == "Barney"|gender == "Female"')
```

```{python filter3-py}
# query with ""; need to use '' for conditions
df.query("name == 'Barney'| gender == 'Female'")
```

</td>
<tr>
</table>




## `group_by()` and `summarize()`

### *Example: Grouped by gender, find mean grade.*

<table class='table'>
<tr> <th>dplyr</th> <th>pandas</th> <tr>
<tr>
<td>

```{r grp-sum1-r}
df %>% 
  group_by(gender) %>% 
  summarize(avg_grade = mean(grade))

```

</td>
<td>

**Option 1:**
```{python grp-sum1a-py}
# returns a series
df.groupby("gender")['grade'].mean()
```

**Option 2:**
```{python grp-sum1b-py}
# returns a data frame
df[['gender', 'grade']].groupby("gender").mean()
```

**Option 3:**
```{python}
# useful for multiple groups and stat
df.groupby(['gender']).agg(
    {'grade': ['mean']}
    # Here key: variable name; value: stat function
)
# here [] is unnecessary but
# required for multiple groups and stats
```

</td>
<tr>
</table>



### *Example: Grouped by gender, find mean, median, minimum, and maximum grade*.


<table class='table'>
<tr> <th>dplyr</th> <th>pandas</th> <tr>
<tr>
<td>


```{r grp-sum2-r}
df %>% 
  group_by(gender) %>% 
  summarize(mean = mean(grade),
            median = median(grade),
            min = min(grade),
            max = max(grade))

```

</td>
<td>
```{python grp-sum2-py}
df.groupby(["gender"]).agg(
  # provide a dictionary
  #   key: variable name
  #   value: stat function
   {'grade':['mean',
             'median',
             'min',
             'max']}
)
```


</td>
<tr>
</table>



## `arrange()`

### *Example: Arrange grade in ascending order.*</span>

<table class='table'>
<tr> <th>dplyr</th> <th>pandas</th> <tr>
<tr>
<td>

```{r arrange-a-r}
df %>% 
  arrange(grade)
```

</td>
<td>
```{python arrange-a-py}
df.sort_values('grade')
```

</td>
<tr>
</table>

### *Example: Arrange grade in descending order.*

<table class='table'>
<tr> <th>dplyr</th> <th>pandas</th> <tr>
<tr>
<td>

```{r arrange-desc-r}
df %>% arrange(desc(grade))
```

</td>
<td>
```{python arrange-desc-py}
df.sort_values('grade', ascending = False)
```

</td>
<tr>
</table>

### *Example: Arrange gender in ascending order then arrange grade in descending order.*

<table class='table'>
<tr> <th>dplyr</th> <th>pandas</th> <tr>
<tr>
<td>

```{r arrange-multiple-r}
df %>%
  arrange(gender, desc(grade))
```

</td>
<td>
```{python arrange-multiple-py}
df.sort_values(['gender','grade'],
                ascending = [True, False])
```
</td>
<tr>
</table>
