# Helper Functions

```{r setup-helper, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r lib-r-helper, echo = FALSE}
library(dplyr)
library(tidyr)
library(reticulate)
```

```{python lib-py-helper, echo = FALSE}
import pandas as pd
import numpy as np
```


## `case_when()` vs `pd.cut()` and `np.select()`

<span style="color:blue">*Suppose we have a data frame with a variable called `age`. We want to create a variable `age_cat` with the following conditions:*

  -  `age` < 18: "Kids"
  
  -  18 $\leq$ `age` < 31: "18-30"
  
  -  31 $\leq$ `age`: "31 and above"</span>


*Example:*

| age |      age_cat |
|----:|-------------:|
|   9 |         Kids |
|  10 |         Kids |
|  18 |        18-30 |
|  21 |        18-30 |
|  29 |        18-30 |
|  31 | 31 and above |
|  45 | 31 and above |


**dplyr::case_when()**

With `dplyr::case_when()` we can do it in the following way:

```{r}
# example data
df <- tibble(age = c(9, 10, 18, 21, 29, 31, 45))
# case_when() in action
df %>% 
  mutate(age_cat = case_when(
    age < 18 ~ "Kids",
    age >= 18 & age < 31 ~ "18-30",
    age >= 31 ~ "31 and above"
  ))
```

We can achieve the same result in Python using 

  - `np.select()`
  
  - `pd.cut()`

```{python}
df = pd.DataFrame({'age': [9, 10, 18, 21, 29, 31, 45]})
```


<table class='table'>
<tr> <th>np.select()</th> <th>pd.cut()</th> <tr>
<tr>
<td>

```{python}
# Step 1: Create conditions
cond = [
  (df['age'].lt(18)),
  (df['age'].ge(18) & (df['age'].lt(31))),
  (df['age'].ge(31))
]

# Step 2: Assign labels
cond_labs = [
  'Kids', '18-30', '30 and above'
]

# Step 3: apply np.select()
df['age_cat'] = np.select(cond, cond_labs)
df
```

</td>
<td>
```{python}
# Step 1: Create bin condition
bin_cond = [0, 17, 30, np.inf] 
# note: instead of 0, 
#       -np.inf will also work
#  
# 0: greater than 0
# 17: upper limit is 17

# Step 2: Assign bin labs
bin_labs = [
  'Kids',
  '18-30',
  '30 and above'
]

# Step 3: apply pd.cut() 
df["age_cat"] = pd.cut(
  df["age"],
  bins = bin_cond,
  labels = bin_labs
)
df
```
</td>
<tr>
</table>

## `if_else()` vs `np.where()`

<span style="color:blue"> *Given prices of shirts `price`, how do we create a variable `price_cat` with the following conditions?*

  - when `price` is less than 50, we label it as "Cheap"
  
  - when `price` is 50 or more, we label it as "Expensive"</span>


<table class='table'>
<tr> <th>dplyr::if_else()</th> <th>np.where()</th> <tr>
<tr>
<td>

```{R}
# toy data
prices <- c(25, 30, 45, 80,100, 125)
df <- tibble(price = prices)
# if_else in action
df %>% 
  mutate(price_cat = if_else(
    price <50, "Cheap", "Expenseive"
  ))
```

</td>
<td>
```{python}
# toy data
prices = {
  'price': [25, 30, 45, 80,100, 125]
}
df = pd.DataFrame(prices)
# np.where() in action
df['price_cat'] = np.where(
  df.price < 50, "Cheap", "Expenseive"
)
df
```
</td>
<tr>
</table>


## `%in%` vs `isin`, `@`, and `in`

|code |capital |
|:----|:-------|
|BD   |Dhaka   |
|PT   |Lisbon  |
|ES   |Madrid  |
|FR   |Paris   |

<span style="color:blue"> *How to keep observations that belong to BD or DE (without using the `|` operator)? *

<table class='table'>
<tr> <th>R</th> <th>Python</th> <tr>
<tr>
<td>

```{R}
# toy data
df <- tibble(
  # country code
  code = c(
    "BD", "PT",
    "ES", "FR"
  ),
  # capital
  capital = c(
    "Dhaka", "Lisbon",
    "Madrid", "Paris"
  ) 
)
df

```

```{r}
# %in% in action
df %>%
  filter(code %in% c("BD", "PT"))
```


</td>
<td>
```{python}
# toy data
data = {
    # country code
    'code':[
        "BD", "PT",
        "ES", "FR"
    ],
    # capital
    'capital':[
        "Dhaka", "Lisbon",
        "Madrid", "Paris"
    ]
}
df = pd.DataFrame(data)
df
```

**isin**

```{python}
# isin in action
df[df["code"].isin(["BD", "PT"])]
```

**`@`**

```{python}
country_list = ["BD", "PT"]
# @ in action
df.query('code == @country_list')
# note: you must create a list first
#   @["BD", "PT"] doesn't work
#   but, @list(["BD", "PT"] works
```
**in**

```{python}
# in in action
df.query('code in ["BD", "PT"]')
```


</td>
<tr>
</table>


## `stringr::str_detect()` vs `str.contains()`

*Example:*

|info             | amount|
|:----------------|------:|
|XYZ Deposit 2020 |      0|
|Cash Deposit     |      1|
|ATM              |      2|
|XYZ Fee 2021     |      3|
|XYZ Deposit 2021 |      4|

<span style="color:blue"> *How to keep or drop only those observations where `info` is about `XYZ`? *

<table class='table'>
<tr> <th>R `stringr::str_detect()` </th> <th>Python `str.contains()`</th> <tr>
<tr>
<td>

```{R}
# toy data
df <- tibble(
  info = c(
    "XYZ Deposit 2020",
    "Cash Deposit",
    "ATM",
    "XYZ Fee 2021",
    "XYZ Deposit 2021"
  ),
  amount = seq(1,5) - 1
)

df

```

**Keep:**

```{r}
# str_detect() in action
df %>%
  filter( # keeps
    stringr::str_detect(
      info, "XYZ"
    )
  )
```

**Drop:**

```{r}
# str_detect() in action
df %>%
  filter( # drops
    ! stringr::str_detect(
      info, "XYZ"
    )
  )
```

</td>
<td>
```{python}
# toy data
df = pd.DataFrame({
    'info':
    [
    "XYZ Deposit 2020",
    "Cash Deposit",
    "ATM",
    "XYZ Fee 2021",
    "XYZ Deposit 2021"
],
    'amount': np.arange(5)    
})
df
```

**Keep:**

```{python}
# str.contains in action: keep
df[df['info'].str.contains("XYZ")]
```

**Drop:**

```{python}
# str.contains in action: drop
df[~ df['info'].str.contains("XYZ")]
```



</td>
<tr>
</table>



## `distinct()` vs `unique()`

|name | year|
|:----|----:|
|A    | 2020|
|B    | 2020|
|C    | 2020|
|A    | 2021|
|B    | 2021|
|D    | 2021|


<span style="color:blue"> *What are the distinct names?*

<table class='table'>
<tr> <th>dplyr::distinct()</th> <th>pd.unique()</th> <tr>
<tr>
<td>

```{R}
# toy data
df <- tibble(
    name = c(
        'A', 'B', 'C',
        'A', 'B', 'D'
    ),
    year = c(
        2020, 2020, 2020,
        2021, 2021, 2021
    )
)
```

```{r}
# dplyr::distinct in action
df %>% distinct(name)
```


```{r}
# if you want to count
df %>% distinct(name) %>% count()
```

</td>
<td>
```{python}
# toy data
df = pd.DataFrame({
    'name':[
        'A', 'B', 'C',
        'A', 'B', 'D'
    ],
    'year':[
        2020, 2020, 2020,
        2021, 2021, 2021
    ]
})

```

```{python}
# pd.unique() in action
df['name'].unique()
```


```{python}
# if you want to count
len(df['name'].unique())
```
</td>
<tr>
</table>


## `slice()` vs `iloc()`

| age|   id|
|---:|----:|
|  20|    1|
|  21|    1|
|  22|    1|
|  23|    1|
|  24|    1|

<span style="color:blue"> *How to slice single or multiple rows?*

<table class='table'>
<tr> <th>dplyr::slice()</th> <th>pd.iloc()</th> <tr>
<tr>
<td>

```{R}
# toy data
df <- tibble(
  age = seq(20,24),
  id = seq(1,5)
)
df
```

```{r}
# slice row 1
df %>% slice(1)
```
```{r}
# slice rows 3 and 4
df %>% slice(3:4)
```
</td>
<td>
```{python}
# toy data
df = pd.DataFrame({
    'age' : np.arange(20, 25),
    'id': np.arange(1, 6)
})
df

```

```{python}
# slice row 1
df.iloc[0:1]
```

```{python}
# slice row 3 and 4
df.iloc[2:4]
```
</td>
<tr>
</table>
