# Join and Bind

## Join Data Frames: `*_join` vs `merge()`

```{r , include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r , echo = FALSE}
library(dplyr)
library(reticulate)
```

```{python , echo = FALSE}
import pandas as pd
import numpy as np
```

**Toy Data**

**`df1`:**

|id   |first_name |
|:----|:----------|
|hiRS |Robin      |
|hiTM |Ted        |
|bbP  |Penny      |
|bbSC |Sheldon    |

**`df2`:**

|id   |last_name  |
|:----|:----------|
|hiRS |Robin      |
|hiTM |Ted        |
|bbSC |Cooper     |
|bbLH |Hofstadter |


<table class='table'>
<tr> <th>R</th> <th>Python</th> <tr>
<tr>
<td>

```{R}
# toy data
df1 = tibble(
    'id'= c("hiRS", "hiTM",
           "bbP", "bbSC"),
    'first_name'= c(
        'Robin',
        'Ted',
        'Penny',
        'Sheldon'
    )
)



df2 = tibble(
    'id'= c("hiRS", "hiTM",
            "bbSC", "bbLH"),
    'last_name'= c(
        'Robin',
        'Ted',
        'Cooper',
        'Hofstadter'
    )
)


```

</td>
<td>
```{python}
# toy data
df1 = pd.DataFrame({
    'id': ["hiRS", "hiTM",
           "bbP", "bbSC"],
    'first_name':[
        'Robin',
        'Ted',
        'Penny',
        'Sheldon'
    ]
})



df2 = pd.DataFrame({
    'id': ["hiRS", "hiTM",
           "bbSC", "bbLH"],
    'last_name':[
        'Scherbatsky',
        'Mosby',
        'Cooper',
        'Hofstadter'
    ]
})

```


</td>
<tr>
</table>

### Inner Join

<table class='table'>
<tr> <th>dplyr::inner_join()</th> <th>pandas::merge(how = "inner")</th> <tr>
<tr>
<td>

```{R}
# inner_join in action
# note: by argument is not needed here
df1 %>% inner_join(df2, by = "id")
```

</td>
<td>
```{python}
# merge(how = "inner") in action
df1.merge(df2, how = 'inner', on = 'id')
# note: default is inner join
# i.e. df1.merge(df2) does inner join
```


</td>
<tr>
</table>



### Left Join

<table class='table'>
<tr> <th>dplyr::inner_join()</th> <th>pandas::merge(how = "left")</th> <tr>
<tr>
<td>

```{R}
# left_join in action
# note: by argument is not needed here
# needed when you have more keys
df1 %>% left_join(df2, by = "id")
```

</td>
<td>
```{python}
# merge(how = "left") in action
df1.merge(df2, how = 'left', on = 'id')
```


</td>
<tr>
</table>



### Right Join

<table class='table'>
<tr> <th>dplyr::right_join()</th> <th>pandas::merge(how = "right")</th> <tr>
<tr>
<td>

```{R}
# left_join in action
# note: by argument is not needed here
# needed when you have more keys
df1 %>% right_join(df2, by = "id")
```

</td>
<td>
```{python}
# merge(how = "right") in action
df1.merge(df2, how = 'right', on = 'id')
```


</td>
<tr>
</table>

### Full Join

<table class='table'>
<tr> <th>dplyr::full_join()</th> <th>pandas::merge(how = "outer")</th> <tr>
<tr>
<td>

```{R}
# full_join in action
# note: by argument is not needed here
# needed when you have more keys
df1 %>% full_join(df2, by = "id")
```

</td>
<td>
```{python}
# merge(how = "outer") in action
df1.merge(df2, how = 'outer', on = 'id')
```


</td>
<tr>
</table>

## Bind Rows: `bind_rows()` vs `append()`

**df1:**

|first_name |last_name |
|:----------|:---------|
|Steve      |Vai       |
|Joe        |Satriani  |


**df2:**

|first_name |last_name |
|:----------|:---------|
|Paul       |Gilbert   |
|Eric       |Johnson   |

**How do we combine the rows of `df1` and `df2`?**

<table class='table'>
<tr> <th>dplyr::bind_rows()</th> <th>pandas::append()</th> <tr>
<tr>
<td>

```{R}
# toy data
df1 <- tibble(
    first_name = c('Steve', 'Joe'),
    last_name = c('Vai', 'Satriani')
)

df2 <- tibble(
    first_name = c('Paul', 'Eric'),
    last_name = c('Gilbert', 'Johnson')
)
```


```{r}
# bind_rows() in action
bind_rows(df1, df2)
```

</td>
<td>
```{python}
# toy data
df1 = pd.DataFrame({
    'first_name': ['Steve', 'Joe'],
    'last_name': ['Vai', 'Satriani']
})

df2 = pd.DataFrame({
    'first_name': ['Paul', 'Eric'],
    'last_name': ['Gilbert', 'Johnson']
})
```

```{python}
# append in action: ignoring the index
df1.append(df2, ignore_index = True)
```


```{python}
# append in action: maintaining the index
df1.append(df2)
```
</td>
<tr>
</table>
