# R Strings
```{r setup-rstring, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
## String Manupulation with Base R Functions


There are many functions in base R for basic string manipulation.


<table class='table'>
<tr> <th>Function</th> <th>Example</th> <tr>
<tr>
<td>

<span style="color:blue">**`nchar()`
**</span>


</td>
<td>

```{r}
y <- c("Hello", "World", "Hello", "Universe")
nchar(y) # Returns number of characters
```

</td>
<tr>

<tr>
<td>

<span style="color:blue">**`tolower()`
**</span>

</td>
<td>

```{r}
tolower(y)
```

</td>
<tr>


<tr>
<td>

<span style="color:blue">**`toupper()`
**</span>

</td>
<td>

```{r}
toupper(y)
```

</td>
<tr>

<tr>
<td>

<span style="color:blue">**`chartr()`
**</span>

</td>
<td>

```{r}
chartr("oe", "$#", y)#o becomes $; e becomes #
```

</td>
<tr>

<tr>
<td>

<span style="color:blue">**`substr()`
**</span>

</td>
<td>

```{r}
x <- "1t345s?"
substr(x, 2, 6) # provides strings from 2 to 6
```

</td>
<tr>
<tr>
<td>

<span style="color:blue">**`strsplit()`
**</span>

</td>
<td>

```{r}
x <- "R#Rocks#!"
strsplit(x, split = "#")
```

</td>
<tr>
</table>

## `stringr`

```{r}
library(stringr)
```

|                   Job                   |     stringr    |   Base R   |
|:---------------------------------------:|:--------------:|:----------:|
|           String concatenation          |    `str_c()`   |  `paste()` |
|           Number of characters          | `str_length()` |  `nchar()` |
|           Extracts substrings           |   `str_sub()`  | `substr()` |
|          Duplicates characters          |   `str_dup()`  |            |
| Removes leading and trailing whitespace |  `str_trim()`  |            |
|              Pads a string              |   `str_pad()`  |            |
|         Wraps a string paragraph        |  `str_wrap()`  |            |

## Regular Expressions

A **regular expression** (or **regex**) is a set of symbols that describes a text pattern. More formally, a regular expression is a pattern that describes a set of strings.


Regular expressions are a formal language in the sense that the symbols have a defined set of rules to specify the desired patterns.

### `stringr` Functions for Regular Expressions

|         Function         |                                     Job                                     |
|:------------------------:|:---------------------------------------------------------------------------:|
| `str_detect`(str, pattern) | Detects the presence of a pattern and returns TRUE if it is found           |
| `str_locate`(str, pattern)          | Locate the 1st position of a pattern and return a matrix with start & end.s |
| `str_extract`(str, pattern)          | Extracts text corresponding to the first match.                             |
| `str_match`(str, pattern)            | Extracts capture groups formed by () from the first match.                  |
| `str_split`(str, pattern)            | Splits string into pieces and returns a list of character vectors.          |
