# Reshape: tidyr vs pandas
```{r setup-tidyr, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r lib-r-tidyr, echo = FALSE}
library(dplyr)
library(tidyr)
library(reticulate)
```

```{python lib-py-tidyr, echo = FALSE}
import pandas as pd
import numpy as np
```

## `pivot_longer()` vs `melt()`

### *Example: Life Expectancy data in "wide" format*

|country    | 1997| 2007|
|:----------|----:|----:|
|Bangladesh | 59.4| 64.1|
|Portugal   | 76.0| 78.1|

<span style="color:blue">*How do we make the table "long"?*</span>

Desired output:

|country    |year | life_exp|
|:----------|:----|--------:|
|Bangladesh |1997 |     59.4|
|Bangladesh |2007 |     64.1|
|Portugal   |1997 |     76.0|
|Portugal   |2007 |     78.1|


<table class='table'>
<tr> <th>tidyr::pivot_longer</th> <th>pandas::melt()</th> <tr>
<tr>
<td>

```{R}
# toy data
df <- tibble(
  country = c("Bangladesh", "Portugal"),
  `1997` = c(59.4, 76.0),
  `2007` = c(64.1, 78.1)
  )
df
```


<span style="color:blue">**`pivot_longer()` in action!**</span>


```
# pivot_longer in action
df %>% 
  pivot_longer(
    cols = c(`1997`, `2007`),
    names_to = "year",
    values_to = "life_exp"
  )
```

</td>
<td>
```{python}
# toy data
data = {
  'country': ["Bangladesh", "Portugal"],
  '1997': [59.4, 76.0],
  '2007': [64.1, 78.1]
}
df = pd.DataFrame(data)
df
```


<span style="color:blue">**`melt()` in action!**</span>

```
# melt() in action
df.melt(
  id_vars = 'country',
  value_vars = ['1997', '2007'], # cols
  var_name = 'year', # names_to
  value_name = 'life_exp' # values_to
)
```

</td>
<tr>
</table>

## `pivot_wider` vs `pivot`

### *Example: Life Expectancy data in "long" format*

|country    |year | life_exp|
|:----------|:----|--------:|
|Bangladesh |1997 |     59.4|
|Bangladesh |2007 |     64.1|
|Portugal   |1997 |     76.0|
|Portugal   |2007 |     78.1|

<span style="color:blue">*How do we make the table "wide"?*</span>

Desired output:

|country    | 1997| 2007|
|:----------|----:|----:|
|Bangladesh | 59.4| 64.1|
|Portugal   | 76.0| 78.1|


<table class='table'>
<tr> <th>tidyr::pivot_wider()</th> <th>pandas::pivot()</th> <tr>
<tr>
<td>

```{R}
# toy data

country <- c(
  "Bangladesh", "Bangladesh",
  "Portugal", "Portugal"
)

year <- c(
  "1997", "2007",
  "1997", "2007"
)

life_exp <- c(
  59.4, 64.1,
  76, 78.1
)

df <- tibble(country, year, life_exp)
df

```

<span style="color:blue">**`pivot_wider()` in action!**</span>


```{r}
# pivot_wider in action
df %>% 
  pivot_wider(
    names_from = "year",
    values_from = "life_exp"
  )
```


</td>
<td>
```{python}
# toy data

country =[
    "Bangladesh", "Bangladesh",
    "Portugal", "Portugal"
]

year = [
    "1997", "2007",
    "1997", "2007"
]

life_exp = [
    59.4, 64.1,
    76, 78.1
]

df = pd.DataFrame(
   {'country': country,
    'year': year,
    'life_exp': life_exp}
)
df
```

<span style="color:blue">**`pivot()` in action!**</span>
```{python}
# pivot in action
df_wide = df.pivot(
    index = 'country',
    columns = 'year', # names from
    values = 'life_exp' # vasles from
)
df_wide
```
```{python}
# Reset the names
df_wide.index.name = None
df_wide.columns.name = None
```

```{python}
df_wide
```

</td>
<tr>
</table>

## pandas:: `stack()` 


```{python}
# toy data
df = pd.DataFrame({
    'year': np.arange(2020,2025),
    'Fall': np.linspace(10,15,5),
    'Spring': np.linspace(1, 5,5)
})
df
```

**How to create MultiIndex series?**

```{python}
# step 1: set year as index
df.set_index('year', inplace = True)
df
```

```{python}
# step 2: apply stack()
df_stacked = df.stack()
df_stacked
```
```{python}
# check type
type(df_stacked)
```

```{python}
# check index
df_stacked.index
```

## pandas:: `unstack()` 

```{python}
# toy data
year = [2010, 2010, 2010, 2020, 2020, 2020]
name = ["X", "Y", "Z", "X", "Y", "Z"]
gender = ["M", "F", "F", "M", "F", "F"]
grade = [10, 10, 20, 20, 12.5, 17.5]
df = pd.DataFrame(
    {
    'year': year,
    'name': name,
    'gender': gender,
    'grade': grade
    }
)
df
```

**Suppose we want to find mean `grade` grouped by `year` and `gender`.**

```{python}
# grouped by year and gender:
# find mean grade
df_stat = df.groupby(
    ['year', 'gender']
).agg({
    'grade': ['mean']
})
df_stat
```

**How to get `F` and `M` as columns?**

  - Just apply `unstack()`
  
```{python}
df_unstacked = df_stat.unstack()
df_unstacked
```
```{python}
# To change the names

# reset index
df_unstacked2 = df_unstacked.reset_index()
# rename
df_unstacked2.columns = ['year', 'M', 'F']
df_unstacked2
```
