# SQL

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r, echo = FALSE}
library(dplyr)
library(tidyr)
library(reticulate)
```

```{python, echo = FALSE}
import pandas as pd
import numpy as np
```

## `CREATE` 

```{sql, eval = FALSE}
DROP TABLE students;

--- create toy table: students
CREATE TABLE students
(student_id CHAR(5) NOT NULL,
name VARCHAR(10),
gender VARCHAR(6),
grade INT,
PRIMARY KEY (student_id));

-- insert data
INSERT INTO students
VALUES 
('SQ001', 'Barney', 'Male', 18),
('SQ002', 'Robin', 'Female', 17),
('SQ007', 'Sheldon', 'Male', 20),
('SQ008', 'Penny', 'Female', 13);
```

|student_id |name    |gender | grade|
|:----------|:-------|:------|-----:|
|S1         |Barney  |Male   |    18|
|S2         |Robin   |Female |    17|
|S7         |Sheldon |Male   |    20|
|S8         |Penny   |Female |    13|

## `UPDATE` 

```{sql, eval = FALSE}
-- Change the grade for Robin to 11
UPDATE students
SET grade = 11 
WHERE student_id = 'S2';
```

|student_id |name    |gender | grade|
|:----------|:-------|:------|-----:|
|S1         |Barney  |Male   |    18|
|S2         |Robin   |Female |    11|
|S7         |Sheldon |Male   |    20|
|S8         |Penny   |Female |    13|

## `DELETE`

```{sql, eval = FALSE}
-- Delete the row for Penny
DELETE FROM students
where student_id = 'S8'
```

|student_id |name    |gender | grade|
|:----------|:-------|:------|-----:|
|S1         |Barney  |Male   |    18|
|S2         |Robin   |Female |    11|
|S7         |Sheldon |Male   |    20|

## Grammar (dplyr vs SQL)

### `select()`

<table class='table'>
<tr> <th>dplyr</th> <th>SQL</th> <tr>
<tr>
<td>

```{R, eval = FALSE}
# all columns
students
```

```{r, eval = FALSE}
# select name and grade
students %>% 
  select(name, gender)
```

</td>
<td>

```{SQL, eval = FALSE}
-- select all columns
SELECT * FROM students
```

```{SQL, eval = FALSE}
-- select name and grade
SELECT name, grade FROM students
```


</td>
<tr>
</table>


### `mutate()`


<table class='table'>
<tr> <th>dplyr</th> <th>SQL</th> <tr>
<tr>
<td>


```{r, eval = FALSE}
# new variable: grade_percent
# expressing grade out of 100
students %>% 
  mutate(
    grade_percent = grade/20 * 100
  )
```

</td>
<td>

```{SQL, eval = FALSE}
-- new variable: grade_percent
-- expressing grade out of 100
SELECT (grade/20.0 * 100)
AS grade_percent
FROM students;
```


</td>
<tr>
</table>


### `filter()`


<table class='table'>
<tr> <th>dplyr</th> <th>SQL</th> <tr>
<tr>
<td>


```{r, eval = FALSE}
# keep if gender is female
# or student is Sheldon
students %>% 
  filter(
    name == 'Sheldon' |
    gender == 'Female'
  )
```


### `summarise()` with `group_by()`


<table class='table'>
<tr> <th>dplyr</th> <th>SQL</th> <tr>
<tr>
<td>


```{r, eval = FALSE}
# grouped by gender,
# find average grade
students %>% 
  group_by(gender) %>% 
  summarise(avg_grade = mean(grade))
```
</td>
<td>

```{SQL, eval = FALSE}
-- grouped by gender,
-- find average grade
SELECT gender, AVG(grade)
AS avg_grade
FROM students
GROUP BY gender;
```


</td>
<tr>
</table>

### `arrange()` 


<table class='table'>
<tr> <th>dplyr</th> <th>SQL</th> <tr>
<tr>
<td>


```{r, eval = FALSE}
# sort name 
# in descending order
students %>% 
  arrange(desc(name))
```
</td>
<td>

```{SQL, eval = FALSE}
-- sort name 
-- in descending order
SELECT *
FROM students
ORDER BY name DESC;
```


</td>
<tr>
</table>
